Liveness-Based Garbage Collection for Lazy Languages

### 2015-06-29 paper/paper.tex
Updated Results section first draft

\section{Results}
To demonstrate the effectiveness of liveness-based garbage collection,
we  have  built a  prototype  consisting  of  an interpreter  for  our
language,  a  liveness  analyzer  and  a copying  collector  that  can
optionally use the results of liveness analysis for marking instead of
reachability. When the collector uses liveness for marking, we call it
a  {\em liveness-based  collector} (LGC),  else we  use the  term {\em
  reachability-based  collector}  (RGC).   The  copying  collector  is
neither incremental nor generational.  As a consequence, any cell that
becomes unreachable or  dead is assuredly collected in  the next round
of garbage collection.

While exploring the activation record of a function, LGC uses the DFAs
at  the latest  program  point traversed  in the function's body.
Thus, if  LGC
is invoked  at the program point   $\pi$,  it would  use  the
liveness  DFAs at  $\pi$ itself for exploring the root set of the
current activation record. For any other activation record in the
stack, say for a function ${\mathit f}$ calling ${\mathit g}$ in the
call-chain,
the DFAs at the point of call to  ${\mathit g}$ would be used.

Let $\dfa{\pi}{x}$  denote the DFA  for the variable  and program
point  pair ($x$,  $\pi$).  We  assume that  there is  a function
$\mathsf{initial}(\dfa{\pi}{x})$  returning the initial  state of
$\dfa{\pi}{x}$.
Considering   a   DFA   as   a   table,
$\dfa{\pi}{x}\mathit{(q,sym)}$  returns the  next  state for  the
state $q$ and the  symbol $\mathit{sym}$, where $\mathit{sym}$ is
\acar\      or     \acdr.       We      shall     also      write
$\dfa{\pi}{x}\mathit{(q,sym)\mathsf{?}}$    for    a    predicate
indicating  whether there  is a  transition from  $q$ on
$\mathit{sym}$.  The LGC action to chase the root variable $x$ at
$\pi$ can be described as  follows: If $\mathscr{L}(\dfa{\pi}{x})$ is
empty,  then  nothing  needs  to  be  done.   Otherwise  we  call
$\mathsf{copy}(\dfa{\pi}{x},\mathsf{initial}(\dfa{\pi}{x}),    x)$
in   Figure~\ref{algo:LGC-main-loop}  and  assign   the  returned
pointer to $x$.
The function $\mathsf{move\_to\_tospace}(x)$  copies the value of
$x$  in the  other semi-space  and returns  the new  address.  It
hides  details such as  returning the  forwarding pointer  if the
value  of $x$  is  already copied,  and  creating the  forwarding
pointer otherwise.

The graphs  in Figure~\ref{fig:memory-usage} show the  number of cells
in the heap over time for RGC and LGC---here time is measured in terms
of the number of cons cells allocated. In addition, they also show the
number of  reachable cells and the  number of cells  that are actually
live\footnote{This   is  statically   approximated  by   our  liveness
  analysis}.  Since  the programs have  different memory requirements,
we have tuned the size of heap for each program to ensure a reasonable
number of  collections. An invocation  of RGC decreases the  number of
cells  in heap  till  it touches  the  curve of  reachable cells.   An
invocation of LGC decreases the number  of heap cells to no lower than
the curve of live cells.

@@

The benchmark programs are drawn from the {\tt no-fib} suite and other
sources and have been  manually converted to  ANF.  All  graphs  except
 {\tt  fibheap}  show  fewer  garbage
collector  invocations for  LGC. The  number  of reachable  cells in
{\tt
  fibheap} increases  initially, comes  down before the  heap becomes
gets
full and then  continues to remains low.  This results in  a single
garbage
collection in both RGC and LGC.  The graphs also show the precision of
our liveness analysis.  For  all programs  except  {\tt nperms} and
{\tt lambda} (Figure~\ref{fig:memory-usage-lambda}),  LGC
manages to collect a good portion of the data that are not live.

@@

%% \begin{figure}[p]
%% \begin{tabular}{@{}c@{}c@{}}
%% \hskip -4mm{\epsfig{file=sudoku.eps, height=4cm, width=6.10cm}}
%% &
%% {\epsfig{file=lcss.eps, height=4cm, width=6.10cm}}
%% \\
%% \hskip -4mm{\epsfig{file=gc_bench.eps, height=4cm, width=6.10cm}}
%%  &
%% {\epsfig{file=nperm.eps, height=4cm, width=6.10cm}}
%% \\
%% \hskip -4mm{\epsfig{file=fibheap.eps, height=4cm, width=6.10cm}}
%% &
%% {\epsfig{file=knightstour.eps, height=4cm, width=6.10cm}}
%% \\
%% \hskip -4mm{\epsfig{file=nqueens.eps, height=4cm, width=6.10cm}}
%% &
%% {\epsfig{file=treejoin.eps, height=4cm, width=6.10cm}}
%% \end{tabular}
%%  \caption{Memory  usage of  programs. The blue and the red curves
%%   indicate the number of cons cells in the active semi-space for RGC
%% and
%%   LGC respectively. The black curve represents the number of reachable
%%   cells and the grey curve represents the number of cells that are
%%   actually live (of which liveness analysis does  a static
%%   approximation). x-axis is the time measured in number of cons-cells
%%   allocated. y-axis is the number of cons-cells.}
%% \label{fig:memory-usage} \figrule
%% \end{figure}
The  increased effectiveness  of LGC  over RGC  is also  shown  in the
tables  in  Figure~\ref{fig:experimental-results}.   The  first  table
provides  statistics  regarding the  analysis  itself.  The number  of
states and the analysis times  are within tolerable limits. Precision
of analysis refers to the percentage of dead cells that is collected by
LGC, averaged over all invocations.
The second table shows garbage  collection statistics for RGC and LGC.
LGC  collects larger garbage  per invocation,  drags cells  for lesser
time and requires a smaller heap size ({\em MinHeap}) for program to
run
in comparison with  RGC.

There are a couple of  issues of concern.  The garbage collection time
is larger in the case of LGC for some programs. The reason is that the
cost of consulting the liveness DFA may outweigh the combined benefits
of  fewer   garbage  collections   and  fewer  markings   per  garbage
collection.   The  other issue  is  illustrated  by  the program  {\tt
  lambda}    (memory   usage  shown   separately    in
Figure~\ref{fig:memory-usage-lambda}).  As can be seen from the table
in Figure~\ref{fig:experimental-results}, the  number of
touched cells\footnote{These  are the  cells which are  visited during
  marking phase, often  more than once due to  sharing.} in this
example is much higher for
LGC.  This  increase is due to  excessive sharing among  heap nodes in
this program.  Note  that a node re-visited because  of sharing is not
explored  any   further  during  a  RGC   collection.   However,  this
curtailment cannot happen in  LGC because of the possibility that the
node,  re-visited in a  different liveness  state, may  mark a  set of
cells different from the earlier visit.
\begin{figure*}[t]
\footnotesize

>>>>

\section{Results}
\comment{To demonstrate the effectiveness of liveness-based garbage collection,
we  have  built a  prototype  consisting  of  an interpreter  for  our
language,  a  liveness  analyzer  and  a copying  collector  that  can
optionally use the results of liveness analysis for marking instead of
reachability. When the collector uses liveness for marking, we call it
a  {\em liveness-based  collector} (LGC),  else we  use the  term {\em
  reachability-based  collector}  (RGC).   The  copying  collector  is
neither incremental nor generational.  As a consequence, any cell that
becomes unreachable or  dead is assuredly collected in  the next round
of garbage collection.}

To test the utility of our method we have implemented a garbage-collector 
for a simple scheme like first order language. It consists of an interpreter, 
liveness analyzer and an interchangeable garbage-collector that can use reachability or liveness. Our benchmark consists of programs from nofib suite and some external programs manually converted to ANF.

\comment{
While exploring the activation record of a function, LGC uses the DFAs
at  the latest  program  point traversed  in the function's body.
Thus, if  LGC
is invoked  at the program point   $\pi$,  it would  use  the
liveness  DFAs at  $\pi$ itself for exploring the root set of the
current activation record. For any other activation record in the
stack, say for a function ${\mathit f}$ calling ${\mathit g}$ in the
call-chain,
the DFAs at the point of call to  ${\mathit g}$ would be used.}

The process of liveness-based garbage collection involves going through 
each activation record on the stack and exploring each variable in the 
current activation record (root set). The liveness of each variable is 
determined using the program point and the variable name. The liveness 
of all variables are stored as DFAs. These DFAs are then encoded as a 
table. A variable is live only if an entry is found in this table.
All cells that are live are copied to the live semi-space. In case of 
\CONS\ cells the \CAR\ and \CDR\ pointers are chased and if they are 
live the copied \CONS\ cell will get the updated addresses of its 
\CAR\ and \CDR\ fields.

\comment{
Let $\dfa{\pi}{x}$  denote the DFA  for the variable  and program
point  pair ($x$,  $\pi$).  We  assume that  there is  a function
$\mathsf{initial}(\dfa{\pi}{x})$  returning the initial  state of
$\dfa{\pi}{x}$.
Considering   a   DFA   as   a   table,
$\dfa{\pi}{x}\mathit{(q,sym)}$  returns the  next  state for  the
state $q$ and the  symbol $\mathit{sym}$, where $\mathit{sym}$ is
\acar\      or     \acdr.       We      shall     also      write
$\dfa{\pi}{x}\mathit{(q,sym)\mathsf{?}}$    for    a    predicate
indicating  whether there  is a  transition from  $q$ on
$\mathit{sym}$.  The LGC action to chase the root variable $x$ at
$\pi$ can be described as  follows: If $\mathscr{L}(\dfa{\pi}{x})$ is
empty,  then  nothing  needs  to  be  done.   Otherwise  we  call
$\mathsf{copy}(\dfa{\pi}{x},\mathsf{initial}(\dfa{\pi}{x}),    x)$
in   Figure~\ref{algo:LGC-main-loop}  and  assign   the  returned
pointer to $x$.
The function $\mathsf{move\_to\_tospace}(x)$  copies the value of
$x$  in the  other semi-space  and returns  the new  address.  It
hides  details such as  returning the  forwarding pointer  if the
value  of $x$  is  already copied,  and  creating the  forwarding
pointer otherwise.}

\subsection{Copying closure using liveness vs Copying closures using reachability}
To compare the effect of using reachability to copy closures instead of liveness we take a couple of programs from our benchmark suite and execute them with 
both. As shown in figure-add-figure-here, the number of garbage collections 
in both cases are same but the time taken varies largely. Both these observations can be explained as follows, since we approximate the liveness of closure to 
the liveness at its creation point, the liveness we get is usually grossly over approximated. Using this liveness may not give too much of a benefit compared to reachability as nearly everything would be live at its point of creation. As mentioned in section-ref using liveness for closures may entail multiple traversals over the same structure consuming more time. Therefore we believe that using reachability for closures is a sweet spot in terms of performance for liveness based garbage collectors.

\comment{
The graphs  in Figure~\ref{fig:memory-usage} show the  number of cells
in the heap over time for RGC and LGC---here time is measured in terms
of the number of cons cells allocated. In addition, they also show the
number of  reachable cells and the  number of cells  that are actually
live\footnote{This   is  statically   approximated  by   our  liveness
  analysis}.  Since  the programs have  different memory requirements,
we have tuned the size of heap for each program to ensure a reasonable
number of  collections. An invocation  of RGC decreases the  number of
cells  in heap  till  it touches  the  curve of  reachable cells.   An
invocation of LGC decreases the number  of heap cells to no lower than
the curve of live cells.

@@

The benchmark programs are drawn from the {\tt no-fib} suite and other
sources and have been  manually converted to  ANF.  All  graphs  except
 {\tt  fibheap}  show  fewer  garbage
collector  invocations for  LGC. The  number  of reachable  cells in
{\tt
  fibheap} increases  initially, comes  down before the  heap becomes
gets
full and then  continues to remains low.  This results in  a single
garbage
collection in both RGC and LGC.  The graphs also show the precision of
our liveness analysis.  For  all programs  except  {\tt nperms} and
{\tt lambda} (Figure~\ref{fig:memory-usage-lambda}),  LGC
manages to collect a good portion of the data that are not live.}

@@

%% \begin{figure}[p]
%% \begin{tabular}{@{}c@{}c@{}}
%% \hskip -4mm{\epsfig{file=sudoku.eps, height=4cm, width=6.10cm}}
%% &
%% {\epsfig{file=lcss.eps, height=4cm, width=6.10cm}}
%% \\
%% \hskip -4mm{\epsfig{file=gc_bench.eps, height=4cm, width=6.10cm}}
%%  &
%% {\epsfig{file=nperm.eps, height=4cm, width=6.10cm}}
%% \\
%% \hskip -4mm{\epsfig{file=fibheap.eps, height=4cm, width=6.10cm}}
%% &
%% {\epsfig{file=knightstour.eps, height=4cm, width=6.10cm}}
%% \\
%% \hskip -4mm{\epsfig{file=nqueens.eps, height=4cm, width=6.10cm}}
%% &
%% {\epsfig{file=treejoin.eps, height=4cm, width=6.10cm}}
%% \end{tabular}
%%  \caption{Memory  usage of  programs. The blue and the red curves
%%   indicate the number of cons cells in the active semi-space for RGC
%% and
%%   LGC respectively. The black curve represents the number of reachable
%%   cells and the grey curve represents the number of cells that are
%%   actually live (of which liveness analysis does  a static
%%   approximation). x-axis is the time measured in number of cons-cells
%%   allocated. y-axis is the number of cons-cells.}
%% \label{fig:memory-usage} \figrule
%% \end{figure}
The table in Figure~\ref{fig:experimental-results} shows that LGC improves the 
garbage collection time and efficiency for most of the programs. 
Even for programs where the time taken is more it is very competitive. 
As the number of garbage collections tends to be very large our graphs show only 
a window which is representative of the behavior for that particular benchmark. 
In all the programs we can see that the curve corresponding to liveness regularly 
dips below the reachability curve.

The only area of concern is the huge gap between the actual liveness and the 
liveness percieved by our collector. In case of LGC for eager languages the 
gap was very narrow and almost touched the actual liveness curve. In case of 
lazy langauges the inefficiency is due to the non-live cells which become part of 
a closure and hence get dragged around. Our experiments have shown that even if 
we use the approximated liveness we do not get too much benefits in terms of non-live
 cells inside closures.
\comment{
The  increased effectiveness  of LGC  over RGC  is also  shown  in the
tables  in  Figure~\ref{fig:experimental-results}.   The  first  table
provides  statistics  regarding the  analysis  itself.  The number  of
states and the analysis times  are within tolerable limits. Precision
of analysis refers to the percentage of dead cells that is collected by
LGC, averaged over all invocations.
The second table shows garbage  collection statistics for RGC and LGC.
LGC  collects larger garbage  per invocation,  drags cells  for lesser
time and requires a smaller heap size ({\em MinHeap}) for program to
run
in comparison with  RGC.

There are a couple of  issues of concern.  The garbage collection time
is larger in the case of LGC for some programs. The reason is that the
cost of consulting the liveness DFA may outweigh the combined benefits
of  fewer   garbage  collections   and  fewer  markings   per  garbage
collection.   The  other issue  is  illustrated  by  the program  {\tt
  lambda}    (memory   usage  shown   separately    in
Figure~\ref{fig:memory-usage-lambda}).  As can be seen from the table
in Figure~\ref{fig:experimental-results}, the  number of
touched cells\footnote{These  are the  cells which are  visited during
  marking phase, often  more than once due to  sharing.} in this
example is much higher for
LGC.  This  increase is due to  excessive sharing among  heap nodes in
this program.  Note  that a node re-visited because  of sharing is not
explored  any   further  during  a  RGC   collection.   However,  this
curtailment cannot happen in  LGC because of the possibility that the
node,  re-visited in a  different liveness  state, may  mark a  set of
cells different from the earlier visit.
}
\begin{figure*}[t]
\footnotesize
